---
title: "Syllabus Economía III (Macroeconomía) Module A AE"
author: "Distance Learning Faculty (FAEDIS)"
date: "2020-2"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 4
bibliography: macro_faedis.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r}

library(tidyverse)
library(knitr)
```

# Professor and meeting hours with students

- __Name__

    + Luis Francisco Gomez Lopez

- __Communication through the virtual classroom__

    + Every Friday between 19:30 hours and 20:30 hours: _Primer Corte 30% > Virtual Meetings > Meeting hours with students_
    
    + Doubts and concerns of the topics of the course, please use the _Forum: Questions or concerns_ that can be accessed at the top of the platform
    
        + In this way, doubts can be shared with all students and all course participants can contribute to the learning process
    
    + Complaints regarding grades, please use the _Mesa de Ayuda_, which can be accessed in the right vertical menu of the platform
    
        + To make any complaints, you are asked to wait first for the grades to be published on the News Bulletin Board and to be made based on the feedbacks that will be made taking into account the evaluation parameters

# Presentation and objectives of the subject

## Description

- Credits: 4
- Type: Theoretical-Practice
- Area of Knowledge: Professional
- Semester: 3
- Block: 1 (July 21 – September 20, 2020) 

## Justification

A comprehensive training of business managers involves knowledge of aspects related to multiple disciplines. One of the disciplines that has greater importance for such training is economics and especially the field of macroeconomics, where it allows you to recognize particularly the economic macro environment in which the different organizations participate.

Therefore, basic knowledge of the economic discipline is essential for business managers to be able to face decision-making processes that involve economic criteria through the understanding of macroeconomic variables, such as the gross domestic product (GDP), the employment rate and unemployment, inflation, public spending, interest rates and exchange rates affect an organization.

## General objective

Introduce and familiarize students with the main macroeconomic concepts using technical tools to understand the macro environment in which an organization operates

# Competencies

- Global competency:

    + Diagnose the macroenvironment of the organization using macroeconomic theory to contribute to the competitive improvement of the company.

- Specific competencies:

    + Identify the macroenvironment of the organization using tools of macroeconomic theory to develop and formulate strategies.
    
    + Interpret the macroeconomic information that affects the organization to make investment and financing decisions.

- Competencies SABE curriculum

    + Skills in handling a second language
    + Ability to work in teams
    + Ability to use the scientific method and logical thinking
    
# Schedule and contents

## Reading schedule

```{r}

tibble(Week = c(1:9, 9),
       Material       = c("[@blanchard_macroeconomics_2017]", 
                          "[@lora_tecnicas_2016]", 
                          rep("[@blanchard_macroeconomics_2017]", times = 7),
                          "[@banco_de_la_republica_informe_2020]"),
       Chapter        = c("1 & 2", "2", "3", "4", "5", "6", "17", "18", "19", "2"),
       `Initial page` = c(26, 1, 70, 90, 112, 134, 372, 392, 414, 31),
       `Final page`   = c(63, 31, 89, 107, 132, 153, 391, 413, 433, 67)) %>%
    mutate(`Number of pages` = (`Final page` - `Initial page`) + 1) %>% 
    kable()
```

## Detailed schedule of virtual meetings and activities

### Primer corte 30%

- July 21

    + Start Block 1

    + __Virtual meeting Presentation Syllabus__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Review the course program in advance, which can be downloaded by following the route _Primer corte 30% > Learning Resources > Support material > Presentations > 000_syllabus_econ_3_macro_2020_2_
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- July 23

    + __Virtual meeting Gross Domestic Product (GDP)__
        + __Meeting time__: 19:00 - 19:45 hours 
        + __Prerequisites__: 
            + Read and review [@blanchard_macroeconomics_2017, Chapters 1 & 2]
            + Have a headset with a built-in microphone to isolate noise from the environment

- July 25

    + __Virtual meeting Inflation__
        + __Meeting time__: 10:00 - 10:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapters 1 & 2]
            + Have a headset with a built-in microphone to isolate noise from the environment

- July 27

    + __Homework 1 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapters 1 & 2]

- July 28

    + __Virtual meeting Indicadores de empleo y desempleo__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:  
            + Read and review [@lora_tecnicas_2016, Chapter 2]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- August 2

    + __Homework 1 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have sent the Homework 1 before 22:00 hours
            
- August 3

    + __Midterm Exam 1 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__: 
            + Read and review [@lora_tecnicas_2016, Chapter 2] and [@blanchard_macroeconomics_2017, Chapter 3]
        
- August 4

    + __Virtual meeting The goods market__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__: 
            + Read and review [@blanchard_macroeconomics_2017, Chapter 3]
            + Have a headset with a built-in microphone to isolate noise from the environment

- August 9

    + __Midterm Exam 1 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__:
            + Have sent the Midterm Exam 1 before 22:00 hours

### Segundo corte 30%

- August 10

    + __Homework 2 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapters 4 & 5]
        
- August 11

    + __Virtual meeting Financial markets I__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 4]
            + Have a headset with a built-in microphone to isolate noise from the environment

- August 18

    + __Virtual meeting Goods and financial markets: IS-LM model__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 5]
            + Have a headset with a built-in microphone to isolate noise from the environment

- August 23

    + __Homework 2 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have sent the Homework 2 before 22:00 hours
        
- August 24

    + __Midterm Exam 2 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__: 
            + Read and review [@blanchard_macroeconomics_2017, Chapter 6]

- August 25

    + __Virtual meeting Financial markets II: The extended IS-LM model__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 6]
            + Have a headset with a built-in microphone to isolate noise from the environment

- August 30

    + __Midterm Exam 2 Closing__
        + __Meeting time__: 22:00 hours
        + __Prerequisites__:
            + Have sent the Midterm Exam 2 before 22:00 hours

### Tercer corte 40%

- August 31

    + __Homework 3 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapters 17 & 18]

- September 1

    + __Virtual meeting Openness in goods and financial markets__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 17]
            + Have a headset with a built-in microphone to isolate noise from the environment

- September 8

    + __Virtual meeting The goods market in an open economy__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 18]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- September 13

    + __Homework 3 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have sent the Homework 3 before 22:00 hours

- September 14

    + __Final Exam Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Study [@blanchard_macroeconomics_2017, Chapters 1-6 & 17-19]
            + Study Chapter [@lora_tecnicas_2016, Chapter 2]
            + Read and review [@banco_de_la_republica_informe_2020, Chapter 2]
            
- September 15

    + __Virtual meeting Output, the interest rate and the exchange rate__
        + __Meeting time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@blanchard_macroeconomics_2017, Chapter 19]
            + Have a headset with a built-in microphone to isolate noise from the environment

- September 17
    
    + __Virtual meeting Informe de la Junta Directiva al Congreso de la República__
        + __Opening time__: 19:00 - 19:45 hours
        + __Prerequisites__:
            + Read and review [@banco_de_la_republica_informe_2020, Chapter 2]
            + Have a headset with a built-in microphone to isolate noise from the environment

- September 20

    + __Final Exam Closing__
        + __Closing time__: 00:00 hours
        + __Prerequisites__:
            + Having answered the Final Exam before 22:00 hours
            
# Qualifiable activities, weights and process in case of no presentation

## Activities and weights

- __Primer corte__: 30%

    + __Homework 1__: 15% 
    + __Midterm Exam 1__: 15% 

- __Segundo corte__: 30%

    + __Homework 2__: 15% 
    + __Midterm Exam 2__: 15%

- __Tercer corte__: 40%

    + __Homework 3__: 15%
    + __Final Exam__: 25 %
    
## Process in case of no presentation

- Students who do not present a __Homework__ due to force majeure or illness, must send the respective excuse scanned through the _Mesa de Ayuda_. If it is not sent, the note will be zero. Once the excuse has been satisfactorily received, the student should __not__ present the activity and the corresponding grade will be equal to the immediately subsequent grade obtained in the __Midterm Exam 1__ or __Midterm Exam 2__ or __Final Exam__ according to the case.

- Students who do not present a __Midterm Exam 1__ or __Midterm Exam 2__ or the __Final Exam__, must send and email to _admon.distancia@unimilitar.edu.co_ two days before or two days after the closing date of these activities. However, this request can be made only once during the course. If it is not requested under these conditions, the grade will be zero.

# Study material

- [@blanchard_macroeconomics_2017, Chapters 1-6 & 17-19]

    + The book can be consulted through the route _http://www.ebooks7-24.com.ezproxy.umng.edu.co/?il=7112 > Login with username and password_ [^1]
    
- [@lora_tecnicas_2016, Chapter 2]

- [@banco_de_la_republica_informe_2020, Chapter 2]

[^1]: Remember that the user to access and read the e-book is the same of your institutional email account, but __without the letter d__ and the password is the one used to check grades

# About plagiarism

- __Acuerdo 02 de 2015 (7 de Mayo): Artículo 68. Escala de Calificaciones Numeral f__ 

_"La comprobación de todo fraude o intento de fraude en cualquier tipo de prueba, acarreará al estudiante y a su(s) colaborador(es), una calificación de CERO PUNTO CERO (0.0) en la prueba respectiva. Como quiera que el fraude o intento de cometerlo es una falta grave, el docente debe informar de este hecho al Director del programa respectivo, quien enviará el informe a la Decanatura de la Facultad, con el fin de iniciar los trámites correspondientes para aplicar la sanción disciplinaria prevista en éste reglamento.”_

- __Acuerdo 05 de 2018 (6 de Junio): Artículo 20 que modifica el Artículo 107 Faltas Disciplinarias Numeral c. Número 6 del Acuerdo 02 de 2015 (7 de Mayo)__

_"Hacer plagio en cualquier tipo de documento o desplegar alguna conducta dolosa que viole el régimen de propiedad intelectual de la Universidad."_

# Study material references
