---
title: "Midterm Exam 1"
author:
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
date: ""
output:
  word_document: 
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE,
                      message   = FALSE, 
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(knitr)
library(readxl)

```

# Labor clasification in Colombia

- The data below were estimated by **Departamento Nacional de Estadística - DANE** using the **Gran encuesta integrada de hogares - GEIH**:

```{r table, ft.align = 'center'}

tibble(Indicator = c('Total Population (TP)',
                     'Working Age Population (WAP)',
                     'Economically Active Population (EAP)',
                     'Employed (E)',
                     'Unemployed (U)',
                     'Visible Unemployment (VU)',
                     'Invisible Unemployment (IU)',
                     'Economically Inactive Population (EIP)')) %>% 
    bind_cols(read_excel(path = '004_anexo_empleo_jun_2020.xlsx',
                         sheet = 2, range = 'IA29:IA36', 
                         col_names = "Individuals")) %>% 
    mutate(Individuals = Individuals*1e3,
           Indicator = factor(Indicator, 
                              levels = c('Total Population (TP)',
                                         'Working Age Population (WAP)',
                                         'Economically Active Population (EAP)',
                                         'Economically Inactive Population (EIP)',
                                         'Employed (E)',
                                         'Unemployed (U)',
                                         'Visible Unemployment (VU)',
                                         'Invisible Unemployment (IU)'))) %>% 
    arrange(Indicator) %>% 
    slice(1:3, 5,7) %>% 
    kable(caption = 'Some labor clasification data in Colombia on June 2020')
```

1. Calculate the Population below 10 or 12 years (Rural or Urban) **(4 points)**

2. Calculate the Economically Inactive Population (EIP) **(4 points)**

3. Calculate the Unemployed (U) **(4 points)**

4. Calculate the Invisible Unemployment (IU) **(3 points)**

# Main indicators of the labor market

Using the information and the results calculated above find:

5. Gross participation rate (GPR) **(3 points)**

6. Labor participation rate (LPR) ("Tasa Global de Participación" in Spanish) **(3 points)**

7. Unemployment rate (UR) **(3 points)**

8. Employment rate (ER) **(3 points)**

# Finding the truth about statements on twitter

- Enter into the Bank of the Republic (Colombia) using the route:

**http://www.banrep.gov.co/** > Estadísticas > Actividad económica, mercado laboral y cuentas financieras > 
4. Empleo y desempleo > Tasas de empleo y desempleo > Descargar y consultar: Total Nacional

- Enter into the link:
<**https://twitter.com/AlvaroUribeVel/status/1025061735554842625**>

9. Point out if the 9.1% value is true based on the Bank of the Republic (Colombia) information. **(3 points)**

10. Point out if in any of the presidential periods of Álvaro Uribe Vélez the *unemployment rate (UR)* was lower than 9.1% at some time. **(3 points)**

11. Point out if the hashtag on the twitter message is valid. **(3 points)**

# The goods market

This exercise is taken from: 

**Oliver Blanchard (2017) Macroeconomics (7 Edition)** > Chapter 3 The Goods Market > Questions and Problems > Exercise 2

- The following equations refer to the goods market of an economy in billions of euros:

$$C_t = 480 + 0.5Y_{tD}$$
$$I_t = 110$$
$$T_t = 70$$
$$G_t = 250$$

12. Solve for the goods market equilibrium  **(5 points)**

13. Find equilibrium disposable income $(Y_{tD})$ **(5 points)**

14. Find equilibrium consumption $(C_t)$ **(4 points)**
