---
title: "Midterm Exam 1"
author:
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
- XXXXXXXX (Only code no name)
date: ''
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  word_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE,
                      message   = FALSE, 
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(knitr)
library(readxl)
library(kableExtra)
library(zoo)
library(tidyquant)
```

# Labor clasification in Colombia

- The data below were estimated by **Departamento Nacional de Estadística - DANE** using the **Gran encuesta integrada de hogares - GEIH**:

```{r table, ft.align = 'center'}

data_labor_col <- tibble(Indicator = c('Total Population (TP)',
                     'Working Age Population (WAP)',
                     'Economically Active Population (EAP)',
                     'Employed (E)',
                     'Unemployed (U)',
                     'Visible Unemployment (VU)',
                     'Invisible Unemployment (IU)',
                     'Economically Inactive Population (EIP)')) %>% 
    bind_cols(read_excel(path = '004_anexo_empleo_jun_2020.xlsx',
                         sheet = 2, range = 'IA29:IA36', 
                         col_names = "Individuals")) %>% 
    mutate(Individuals = Individuals*1e3,
           Indicator = factor(Indicator, 
                              levels = c('Total Population (TP)',
                                         'Working Age Population (WAP)',
                                         'Economically Active Population (EAP)',
                                         'Economically Inactive Population (EIP)',
                                         'Employed (E)',
                                         'Unemployed (U)',
                                         'Visible Unemployment (VU)',
                                         'Invisible Unemployment (IU)'))) %>% 
    arrange(Indicator) %>% 
    slice(1:3, 5,7)

data_labor_col %>% 
    kable(caption = 'Some labor clasification data in Colombia on June 2020')  %>% 
    kable_styling(bootstrap_options = 'striped', full_width = FALSE)
```

1. Calculate the Population below 10 or 12 years (Rural or Urban) **(4 points)**

2. Calculate the Economically Inactive Population (EIP) **(4 points)**

3. Calculate the Unemployed (U) **(4 points)**

4. Calculate the Invisible Unemployment (IU) **(3 points)**

```{r}

# Solution
data_sol1 <- data_labor_col %>%
    set_names(nm = c("indicator", "value")) %>% 
    mutate(indicator = c("TP", "WAP", "EAP", "E", "VU")) %>% 
    pivot_wider(names_from = indicator, values_from = value) %>% 
    mutate(rural_urban = TP - WAP,
           EIP         = WAP - EAP,
           U           = EAP - E,
           IU          = U - VU)
  
    # Table
    data_sol1 %>% 
    select(rural_urban: IU) %>% 
    set_names(nm = c("Population below 10 or 12 years (Rural or Urban)",
                     "Economically Inactive Population (EIP)",
                     "Unemployed (U)",
                     "Invisible Unemployment (IU)")) %>% 
    pivot_longer(1:4, names_to = "Indicator", values_to = "Individuals") %>% 
    kable()  %>% 
    kable_styling(bootstrap_options = 'striped', full_width = FALSE)
    
```

# Main indicators of the labor market

Using the information and the results calculated above find:

5. Gross participation rate (GPR) **(3 points)**

6. Labor participation rate (LPR) ("Tasa Global de Participación" in Spanish) **(3 points)**

7. Unemployment rate (UR) **(3 points)**

8. Employment rate (ER) **(3 points)**

```{r}

# Solution
data_sol2 <- data_sol1 %>% 
  mutate(GPR = EAP / TP,
         LPR = EAP / WAP,
         UR  = U / EAP,
         ER  = E / WAP) %>% 
  select(GPR:ER) %>% 
    set_names(nm = c("Gross participation rate (GPR)",
                     "Labor participation rate (LPR)",
                     "Unemployment rate (UR)",
                     "Employment rate (ER)")) %>% 
    pivot_longer(1:4, names_to = "Indicator", values_to = "Rates")
    
    # Table
    data_sol2 %>% 
    mutate(Rates = scales::percent(Rates)) %>% 
    kable()  %>% 
    kable_styling(bootstrap_options = 'striped', full_width = FALSE)
```

# Finding the truth about statements on twitter

- Enter into the Bank of the Republic (Colombia) using the route:

**http://www.banrep.gov.co/** > Estadísticas > Actividad económica, mercado laboral y cuentas financieras > 
4. Empleo y desempleo > Tasas de empleo y desempleo > Descargar y consultar: Total Nacional

- Enter into the link:
<**https://twitter.com/AlvaroUribeVel/status/1025061735554842625**>

9. Point out if the 9.1% value is true based on the Bank of the Republic (Colombia) information. **(3 points)**

The value of 9.1% is true. Based on the Bank of the Republic (Colombia) information, the unemployment rate was 9.08% in June, 2018 so this value was rounded and pointed out on twitter.

10. Point out if in any of the presidential periods of Álvaro Uribe Vélez the *unemployment rate (UR)* was lower than 9.1% at some time. **(3 points)**

In none of the presidential terms of Alvaro Uribe Velez, unemployment was less than 9.1%. You can see that in the plot below.

```{r, fig.width=10, fig.height=6}

# Data 1: presidents Colombia
presi_col <- read_excel(path = "004_presi_col.xlsx") %>% 
                slice(2:5) %>% 
                mutate(start      = ymd(start),
                       end        = ymd(end),
                       name_label = str_glue('{word(name,-2)} {word(name,-1)}') %>% 
                                 fct_reorder(start)) 

# Data 2: unemployment rate
ur_col <- read_excel("004_1.1.1.EMP_Total_nacional_IQY.xlsx", 
           sheet     = 1, 
           range     = "A11:C243", 
           col_names = c("date", "ER", "UR")) %>% 
  mutate(ER = ER / 100,
         UR = UR / 100) %>% 
  mutate(date = as.yearmon(date, "%Y-%m") %>% 
           as.Date(frac = 1)) %>% 
  select(date, UR)

  # Plot
  ur_col %>% 
    ggplot() +
    geom_point(aes(x = date, y = UR), 
               shape = 21, 
               fill = palette_light()[[2]],
               color = "black") + 
    geom_line(aes(x = date, y = UR)) +
    geom_hline(yintercept = 0.0908,
               color      = palette_light()[[3]]) + 
    geom_rect(data  = presi_col,
              aes(xmin  = start, xmax = end, fill = name_label), 
              ymin  = -Inf, 
              ymax  = Inf, 
              alpha = 0.3,
              color = "black",
              show.legend = TRUE) +
    scale_y_continuous(breaks = c(0, 0.05, 0.0908, 0.15, 0.20), 
                     labels = scales::percent_format(accuracy = 0.01)) +
    scale_fill_tq() +
    scale_x_date(breaks = c(ymd('2001-01-31'),
                            ymd('2005-01-31'),
                            ymd('2010-01-31'),
                            ymd('2015-01-31'),
                            ymd('2020-01-31')),
                 date_labels = '%Y-%m') +
  labs(x        = '',
       y        = 'Percent',
       fill     = 'Presidential period',
       title    = 'Total national unemployment rate (UR) Colombia',
       caption  = 'Source: Banco de la República - Colombia')  +
    theme(panel.border      = element_rect(fill = NA, color = "black"),
          plot.background   = element_rect(fill = "#f3fcfc"),
          panel.background  = element_rect(fill = "#f3f7fc"),
          legend.background = element_rect(fill = "#f3fcfc"),
          plot.title        = element_text(face = "bold"),
          axis.title        = element_text(face = "bold"),
          legend.title      = element_text(face = "bold"),
          legend.position   = "bottom", 
          axis.text         = element_text(face = "bold"),
          axis.text.x = element_text(angle = 90, vjust = 0.5))
    
```


11. Point out if the hashtag on the twitter message is valid. **(3 points)**

The hashtag is not valid and is also quite misleading. If it were true the country would be broken long ago.

# The goods market

This exercise is taken from: 

**Oliver Blanchard (2017) Macroeconomics (7 Edition)** > Chapter 3 The Goods Market > Questions and Problems > Exercise 2

- The following equations refer to the goods market of an economy in billions of euros:

$$C_t = 480 + 0.5Y_{tD}$$
$$I_t = 110$$
$$T_t = 70$$
$$G_t = 250$$

12. Solve for the goods market equilibrium  **(5 points)**

- Goods market equilibrium  
 
 $$Y_t = Z_t$$
We are assuming that the economy is closed, that is, there are no commercial relationships with the rest of the world:

$$\begin{split}
  Y_t    & = C_t + I_t + G_t \\
  Y_t    & = 480 + 0.5Y_{tD} + I_t + G_t \\
  Y_t    & = 480 + 0.5(Y_t - T_t) + I_t + G_t \\
  0.5Y_t & = 480 - 0.5T_t + I_t + G_t \\
  Y_t    & = \frac{480 - 0.5T_t + I_t + G_t}{0.5} \\
  Y_t    & = \frac{480 - (0.5*70) + 110 + 250}{0.5} \\
  Y_t    & = \frac{805}{0.5} \\
  Y_t    & = 1610
  \end{split}$$

13. Find equilibrium disposable income $(Y_{tD})$ **(5 points)**

$$\begin{split}
   Y_{tD} & = Y_t - T_t \\
   Y_{tD} & = 1610 - 70 \\
   Y_{tD} & = 1540
   \end{split}$$

14. Find equilibrium consumption $(C_t)$ **(4 points)**

$$\begin{split}
   C_t = 480 + 0.5Y_{tD} \\
   C_t = 480 + 0.5*1540 \\
   C_t = 1250
   \end{split}$$
