---
title: 'The Goods Market in an Open Economy'
author: "Luis Francisco Gomez Lopez"
date: 2020-09-06
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
  ioslides_presentation: default
  slidy_presentation: default
bibliography: macro_faedis.bib
link-citations: yes
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r}

library(tidyverse)
library(latex2exp)
library(readxl)
library(knitr)
library(kableExtra)
```

# Contents

- Please Read Me
- Purpose
- Demand for goods in an open economy
- IS curve in an open economy
- Marshall–Lerner condition
- IS curve and the Balance of Payments (BOP)
- Acknowledgments
- References

# Please Read Me

- Check the message __Welcome greeting__ published in the News Bulletin Board.

- Dear student please edit your profile uploading a photo where your face is clearly visible.

- The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

- This presentation is based on [@blanchardMacroeconomics2017a, Chapter 18]

# Purpose

Analyze the equilibrium of the goods market of an open economy.

# Demand for goods in an open economy

 - The demand of goods in an open economy is:
 
    + $\widehat{Z} \equiv \widehat{C} + \widehat{I} + \widehat{G} + \widehat{X} - \varepsilon*\widehat{IM}$
    
    + If we include behavioral equations we have[^1]: $$\widehat{Z} = \widehat{C}(\widehat{Y}-\widehat{T}) + \widehat{I}(\widehat{Y}, r) + \widehat{G} + \widehat{X}(\widehat{Y}^*,\varepsilon) - \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$$
     
        + Where exports and imports, $\widehat{X}$ and $\widehat{IM}$, depend on the **real multilateral** exchange rate, $\varepsilon$. Also exports depend on **GDP** inside the territory, $\widehat{Y}$, and imports depend on **GDP** outside the territory, $\widehat{Y}^*$
        
        + Furthermore, we multiply $\widehat{IM}$ by $\varepsilon$ to express the term $\varepsilon*\widehat{IM}$ in the domestic currency[^2].

[^1]: We are not going to include the **risk premium**, $x$, to facilitate the analysis.
[^2]: If the different **nominal** exchange rates, that are included in $\varepsilon$, are expressed as the amount of **units of national currency** that must be given in exchange for a **unit of foreign currency** then you have to multiply and not divide the imports by $\varepsilon$.

# Demand for goods in an open economy

 - The demand of goods in an open economy is:
 
     + $\widehat{Z} \equiv \widehat{C} + \widehat{I} + \widehat{G} + \widehat{X} - \varepsilon*\widehat{IM}$
     
         + If we include behavioral equations we have: $$\widehat{Z} = \widehat{C}(\widehat{Y}-\widehat{T}) + \widehat{I}(\widehat{Y}, r) + \widehat{G} + \widehat{X}(\widehat{Y}^*,\varepsilon) - \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$$

    + $\widehat{C}$, $\widehat{I}$ and $\widehat{G}$ include both products that are produced within the economy and products that are produced in the rest of the world.

        + Therefore these variables can depend on $\varepsilon$. In that sense, we assume that if $\varepsilon$ increases, the products produced by the **rest of the world** are replaced by products produced **within** the economy and together $\widehat{C}$, $\widehat{I}$ and $\widehat{G}$ do not change.

# Demand for goods in an open economy

- The demand for goods in an open economy is divided between:

    + Demand for goods **within the territory** that can be produced **within the territory** or in the **rest of the world**: $\widehat{C}(\widehat{Y}-\widehat{T}) + \widehat{I}(\widehat{Y}, r) + \widehat{G}$

    + Demand for goods from the **rest of the world** produced **within the territory**: $\widehat{X}(\widehat{Y}^*,\varepsilon)$

    + Demand for goods **within the territory** produced in the **rest of the world**: $\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$

- Additionally it is important to mention that for a given period it can happen that:

    + $\widehat{X}(\widehat{Y}^*,\varepsilon) > \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$ (**Trade surplus**)
    
    + $\widehat{X}(\widehat{Y}^*,\varepsilon) < \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$ (**Trade deficit**)
    
    + $\widehat{X}(\widehat{Y}^*,\varepsilon) = \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$

# Demand for goods in an open economy

```{r, out.width = '90%'}

ggplot(data = tibble(x = c(0, 5,  0,  5),
                     y = c(5, 0, -1, -6))
) +
    #First plot
    geom_point(aes(x = x, y = y),
               color = "blue",
               size  = 3) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 5, yend = 0)) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 0, yend = 5)) + 
    geom_segment(aes(x = 0, y = 1.5, 
                     xend = 4, yend = 3.5,
    ),
    color = 'red') +
    annotate(geom = 'text', 
             x = 4, y = 3.75, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G}')) +
    annotate(geom = 'text', 
             x = 5.25, y = 0, 
             label = TeX('$\\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 0, y = 5.25, 
             label = TeX('$\\widehat{Z}')) +
    
    #Second plot
    geom_segment(aes(x = 0, y = -6, 
                     xend = 0, yend = -1)) +
    geom_segment(aes(x = 0, y = -6, 
                     xend = 5, yend = -6)) +
    geom_segment(aes(x = 0, y = -4.5, 
                     xend = 4, yend = -2.5,
    ),
    color = 'red') +
    geom_segment(aes(x = 0, y = -5, 
                     xend = 4, yend = -4,
    ),
    color = 'green') +
    annotate(geom = 'text', 
             x = 5.25, y = -6, 
             label = TeX('$\\widehat{Y}$')) +
    annotate(geom = 'text', 
             x = 0, y = -0.75, 
             label = TeX('$\\widehat{Z}')) + 
    annotate(geom = 'text', 
             x = 4, y = -2.25, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G}')) +
    annotate(geom = 'text', 
             x = 4, y = -3.75, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G} - \\epsilon\\widehat{M}(\\widehat{Y}_R,\\epsilon)')) + 
    theme_void()
```

# Demand for goods in an open economy

```{r, out.width = '90%'}

ggplot(data = tibble(x = c(0, 5,  0,  5,  0,  2, 2),
                     y = c(5, 0, -1, -4, -6, -4, 2.5))
) +
    #First plot
    geom_point(aes(x = x, y = y),
               color = "blue",
               size  = 3) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 5, yend = 0)) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 0, yend = 5)) + 
    geom_segment(aes(x = 0, y = 1.5, 
                     xend = 4, yend = 3.5,
    ),
    color = 'red') +
    geom_segment(aes(x = 0, y = 1, 
                     xend = 4, yend = 2,
    ),
    color = 'green') +
    geom_segment(aes(x = 0, y = 2, 
                     xend = 4, yend = 3,
    ),
    color = 'purple') +
    annotate(geom = 'text', 
             x = 4, y = 3.75, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G}')) +
    annotate(geom = 'text', 
             x = 4, y = 2.25, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G} - \\epsilon\\widehat{M}(\\widehat{Y}_R,\\epsilon)')) +
    annotate(geom = 'text', 
             x = 4, y = 3.25, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G} - \\epsilon\\widehat{IM}(\\widehat{Y}_R,\\epsilon) + \\widehat{X}(\\widehat{Y}^*,\\epsilon)$')) +
    annotate(geom = 'text', 
             x = 5.25, y = 0, 
             label = TeX('$\\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 0, y = 5.25, 
             label = TeX('$\\widehat{Z}')) +
    
    #Second plot
    geom_segment(aes(x = 0, y = -6, 
                     xend = 0, yend = -1)) +
    geom_segment(aes(x = 0, y = -4, 
                     xend = 5, yend = -4)) +
    geom_segment(aes(x = 0, y = -2, 
                     xend = 4, yend = -6),
                 color = 'orange') +
    geom_segment(aes(x = 2, y = 2.5, 
                     xend = 2, yend = -4),
                 linetype = 'dashed') +
    annotate(geom = 'text', 
             x = 5.25, y = -4, 
             label = TeX('$\\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 0.25, y = -0.75, 
             label = TeX('$\\widehat{X}(\\widehat{Y}^*,\\epsilon) - \\epsilon\\widehat{IM}(\\widehat{Y}_R,\\epsilon)$')) +
    annotate(geom = 'text', 
             x = 0.5, y = -3.5, 
             label = 'Trade surplus') +
    annotate(geom = 'text', 
             x = 3.25, y = -4.5, 
             label = 'Trade deficit') + 
    annotate(geom = 'text', 
             x = -0.1, y = -4, 
             label = '0') +
    theme_void()
```

# IS curve in an open economy

- *IS* curve:

    + $\widehat{Y} = \widehat{Z}$

    + $\widehat{Y} = \widehat{C}(\widehat{Y}_R-\widehat{T}) + \widehat{I}(r,\widehat{Y}) + \widehat{G} + \widehat{X}(\widehat{Y}^*,\varepsilon) - \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$
    
- The **equilibrium condition** in the Goods Market for an Open Economy can by achieve with:

    + $\widehat{X}(\widehat{Y}^*,\varepsilon) > \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$
    
    + $\widehat{X}(\widehat{Y}^*,\varepsilon) < \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$
    
    + $\widehat{X}(\widehat{Y}^*,\varepsilon) = \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$
    
        + With a **trade surplus**, **trade deficit** or **without them** the Goods Market can be in equilibrium.
    
# IS curve in an open economy

- **Example** of the **equilibrium condition** in the Goods Market with $\widehat{X}(\widehat{Y}^*,\varepsilon) < \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$ (**Trade deficit**)[^3]

```{r, out.width = '65%'}

ggplot(data = tibble(x = c(0, 5,  0,  5,  0,  2, 8/3,  8/3),
                     y = c(5, 0, -1, -4, -6, -4, 8/3, -14/3))
) +
    #First plot
    geom_point(aes(x = x, y = y),
               color = "blue",
               size  = 3) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 5, yend = 0)) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 0, yend = 5)) +
    geom_segment(aes(x = 0, y = 0, 
                     xend = 4, yend = 4)) +
    geom_segment(aes(x = 0, y = 2, 
                     xend = 4, yend = 3,
    ),
    color = 'purple') +
    annotate(geom = 'text', 
             x = 4, y = 4.25, 
             label = TeX('$\\widehat{Y} \\equiv \\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 4, y = 3.25, 
             label = TeX('$\\widehat{Z} = \\widehat{C}(\\widehat{Y}_R-\\widehat{T}) + \\widehat{I}(r,\\widehat{Y}_R) + \\widehat{G} - \\epsilon\\widehat{IM}(\\widehat{Y}_R,\\epsilon) + \\widehat{X}(\\widehat{Y}^*,\\epsilon)$')) +
    annotate(geom = 'text', 
             x = 5.25, y = 0, 
             label = TeX('$\\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 0, y = 5.25, 
             label = TeX('$\\widehat{Z}\\;\\widehat{Y}')) +
    
    #Second plot
    geom_segment(aes(x = 0, y = -6, 
                     xend = 0, yend = -1)) +
    geom_segment(aes(x = 0, y = -4, 
                     xend = 5, yend = -4)) +
    geom_segment(aes(x = 0, y = -2, 
                     xend = 4, yend = -6),
                 color = 'orange') +
    geom_segment(aes(x = 8/3, y = 8/3, 
                     xend = 8/3, yend = -14/3),
                 linetype = 'dashed') +
    annotate(geom = 'text', 
             x = 5.25, y = -4, 
             label = TeX('$\\widehat{Y}_R$')) +
    annotate(geom = 'text', 
             x = 0.25, y = -0.75, 
             label = TeX('$\\widehat{X}(\\widehat{Y}^*,\\epsilon) - \\epsilon\\widehat{IM}(\\widehat{Y}_R,\\epsilon)$')) +
    annotate(geom = 'text', 
             x = 0.5, y = -3.5, 
             label = 'Trade surplus') +
    annotate(geom = 'text', 
             x = 3.25, y = -4.5, 
             label = 'Trade deficit') + 
    annotate(geom = 'text', 
             x = -0.1, y = -4, 
             label = '0') +
    theme_void()
```

[^3]: Remember that it can exist equilibriums where $\widehat{X}(\widehat{Y}^*,\varepsilon) \geq \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$

# Marshall–Lerner condition

- The **real** balance trade or **real** net exports, is defined as $\widehat{NX} \equiv \widehat{X} - \widehat{IM}$

    + Using the **behavioral equations** we have that $\widehat{NX} = \widehat{X}(\widehat{Y}^*,\varepsilon) - \varepsilon\widehat{IM}(\widehat{Y},\varepsilon)$

- A natural question that arises is how $\widehat{NX}$ depends on the **real multilateral** exchange rate, $\varepsilon$. If that question can be answered, we can also determine how $\varepsilon$ is related with the **real** GDP, $\widehat{Y}$.

- To answer this question we need to use the concept of **derivative** from differential calculus and the concept of **elasticity** from microeconomics.[^4]

[^4]: If you don't have these tools, please skip this section and focus on the numerical example at the end of the explanation

# Marshall–Lerner condition

- To know the relationship between $\widehat{NX}$ and $\varepsilon$ we must know what determines the sign of $\frac{d\widehat{NX}}{d\varepsilon}$:

\footnotesize
$$\begin{split}
   \frac{d\widehat{NX}}{d\varepsilon} & = \frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon} - \frac{d\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}{d\varepsilon} \\
   & = \frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon} - \left[\widehat{IM}(\widehat{Y},\varepsilon) + \varepsilon\frac{d\widehat{IM}(\widehat{Y},\epsilon)}{d\varepsilon}\right] \\
   & = \frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon} - \widehat{IM}(\widehat{Y},\varepsilon)\left[1 + \frac{d\widehat{IM}(\widehat{Y},\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{IM}(\widehat{Y},\varepsilon)}\right] \\
   & = \frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon}\frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{X}(\widehat{Y}^*,\varepsilon)} - \widehat{IM}(\widehat{Y},\varepsilon)\left[1 + \frac{d\widehat{IM}(\widehat{Y},\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{IM}(\widehat{Y},\varepsilon)}\right]
   \end{split}$$

# Marshall–Lerner condition

- To know the relationship between $\widehat{NX}$ and $\varepsilon$ we must know what determines the sign of $\frac{d\widehat{NX}}{d\varepsilon}$:

\footnotesize
$$\begin{split}
   \frac{d\widehat{NX}}{d\varepsilon} & =
   \widehat{IM}(\widehat{Y},\varepsilon)\left[\frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{X}(\widehat{Y}^*,\varepsilon)} -  \frac{d\widehat{IM}(\widehat{Y},\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{IM}(\widehat{Y},\varepsilon)} - 1\right] \\
   & = \widehat{IM}(\widehat{Y},\varepsilon)\left[\frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} - 1\right]
   \end{split}$$

Where $\eta_{\widehat{X},\varepsilon} \equiv \frac{d\widehat{X}(\widehat{Y}^*,\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{X}(\widehat{Y}^*,\varepsilon)}$ and $\eta_{\widehat{IM},\varepsilon} \equiv \frac{d\widehat{IM}(\widehat{Y},\varepsilon)}{d\varepsilon}\frac{\varepsilon}{\widehat{IM}(\widehat{Y},\varepsilon)}$

# Marshall–Lerner condition

- $\eta_{\widehat{X},\varepsilon}$ is the elasticity of the **real multilateral** exchange rate with respect to the **real** exports and $\eta_{\widehat{IM},\varepsilon}$ is the elasticity of the **real multilateral** exchange rate with respect to the **real** imports.

- The sign of $\frac{d\widehat{NX}}{d\varepsilon}$ is determined by:

$$\frac{d\widehat{NX}}{d\varepsilon} =
\left\{
	\begin{array}{ll}
		>0  & \mbox{if } \frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} > 1 \\
		=0 & \mbox{if }  \frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} = 1 \\
		<0 & \mbox{if }  \frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} < 1
	\end{array}
\right.$$

- If $\frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} > 1$ then $\frac{d\widehat{NX}}{d\varepsilon} > 0$ and this situation is known as the **Marshall–Lerner condition**. Also if this condition is fulfilled then there is a positive relation between $\varepsilon$ and $\widehat{Y}$. 

# Marshall–Lerner condition

- Numerical example of the **Marshall–Lerner condition**

    + Let us assume that $\varepsilon$ increases by 1% and we want to know how much $\widehat{X}$ and $\widehat{IM}$ decrease or increase in percentage terms. 
    
        + The concept of **elasticity** allows us to answer the previous question where it indicates that happens to a dependent variable in percentage terms when an independent variable increases by 1% starting from some initial values of the dependent and independent variable.
        
    + Let us assume that $\widehat{X}(\widehat{Y}^*,\varepsilon) = \varepsilon\widehat{IM}(\widehat{Y},\varepsilon) > 0$ so $\frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)} = 1$. 
    
    + Also assume that $\eta_{\widehat{X},\varepsilon} = 0.9$ and $\eta_{\widehat{IM},\varepsilon} = -0.8$. This means that if $\varepsilon$ increases by 1% then $\widehat{X}$ increases by 0.9% and $\widehat{IM}$ decreases by -0.8%.
    
        + In this case $\frac{\widehat{X}(\widehat{Y}^*,\varepsilon)}{\varepsilon\widehat{IM}(\widehat{Y},\varepsilon)}\eta_{\widehat{X},\varepsilon} - \eta_{\widehat{IM},\varepsilon} = 1*0.9 - (-0.8) = 1.7 > 1$. Therefore, the **Marshall–Lerner condition** is fulfilled for this particular **numerical example**, $\frac{d\widehat{NX}}{d\varepsilon} > 0$.

# IS curve and the Balance of Payments (BOP)

- The **IS** curve for an open economy represents the equilibrium in the Goods Market:

\footnotesize
$$\begin{split}
   \widehat{Y} & = \widehat{C}(\widehat{Y}_R-\widehat{T}) + \widehat{I}(r,\widehat{Y}) + \widehat{G} + \widehat{X}(\widehat{Y}^*,\varepsilon) - \varepsilon\widehat{IM}(\widehat{Y},\varepsilon) \\
   \widehat{Y} & = \widehat{C}(\widehat{Y}_R-\widehat{T}) + \widehat{I}(r,\widehat{Y}) + \widehat{G} + \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) \\
   \widehat{Y} - \widehat{C}(\widehat{Y}_R - \widehat{T}) & = \widehat{I}(r,\widehat{Y}) + \widehat{G} + \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) \\
   \widehat{Y} - \widehat{C}(\widehat{Y}_R - \widehat{T}) - \widehat{T} & = \widehat{I}(r,\widehat{Y}) - (\widehat{T} - \widehat{G}) + \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon)
   \end{split}$$
    
- In an open economy the income of individuals **within** a territory includes the income obtained **within** the territory and represented by $\widehat{Y}$ as well as other income from the **rest of the world**. 

- We are going to include this aspect using elements from the **Balance of Payments (BOP)** and taking into account [@internationalmonetaryfundBalancePaymentsInternational2009]
    
# IS curve and the Balance of Payments (BOP)

\tiny
```{r}

read_excel(path  = '009_bp_resumen_IQY.xlsx',
           sheet = 1,
           range = 'A11:V53') %>%
  select(1:2) %>%
  set_names(nm = c('Account', 'Value (Millons USD)')) %>%
  slice(1:6, 13:21, 24, 27:28, 31, 34:41) %>%
  kable(format = 'latex',
        digits = 2, caption = 'Balance of Payments (BOP) for Colombia in 2000') %>%
  kable_styling(latex_options = 'scale_down') %>%
  row_spec(c(1, 13, 27), bold = TRUE, color = 'white', background = '#e31a1c') %>%
  row_spec(c(4, 7, 10, 14, 17, 20, 23, 26), bold = TRUE, color = 'white', background = '#18BC9C') %>%
  footnote(general           = 'Banco de la República - Colombia',
           general_title     = 'Source: ',
           number            = 'Methodology: Sixth version of the Balance of Payments Manual of the International Monetary Fund (IMF)',
           alphabet = 'The Capital account does not appear because the sources of information currently available do not allow the identification and registration of capital transfers for Colombia',
           footnote_as_chunk = TRUE)
```

# IS curve and the Balance of Payments (BOP)

- The **Primary Income** (**Ingreso primario (Renta factorial)**) and the **Secondary Income** (**Ingreso secundario (Transferencias corrientes)**) represents the other income from the **rest of the world**.

- If the **Primary Income** (**Ingreso primario (Renta factorial)**) is represented by $\widehat{NI}$ and the **Secondary Income** (**Ingreso secundario (Transferencias corrientes)**) is represented by $\widehat{NT}$ we can rewrite the **IS** curve for an open economy as:

\footnotesize
$$\begin{split}
  (\widehat{Y} + \widehat{NI} + \widehat{NT} - \widehat{T}) - \widehat{C}(\widehat{Y}_R - \widehat{T}) & = \widehat{I}(r,\widehat{Y}) - (\widehat{T} - \widehat{G}) + \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) + \widehat{NI} + \widehat{NT} \\
  \widehat{S}^{pr} & = \widehat{I}(r,\widehat{Y}) - \widehat{S}^{pu} + \widehat{CA} \\
  \widehat{S}^{pr} + \widehat{S}^{pu} - \widehat{I}(r,\widehat{Y}) & = \widehat{CA}
  \end{split}$$
  
 - Where $\widehat{S}^{pr} \equiv (\widehat{Y} + \widehat{NI} + \widehat{NT} - \widehat{T}) - \widehat{C}(\widehat{Y}_R - \widehat{T})$  is the **real** private savings in an open economy, $\widehat{S}^{pu} \equiv \widehat{T} - \widehat{G}$ is the **real** public savings in an open economy and $\widehat{CA} \equiv \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) + \widehat{NI} + \widehat{NT}$ is the **current account**.

# IS curve and the Balance of Payments (BOP)
        
- In that sense, if the Goods Market is in equilibrium then the difference between savings and investment is equal to the **current account**.

\footnotesize
$$\begin{split}
  (\widehat{Y} + \widehat{NI} + \widehat{NT} - \widehat{T}) - \widehat{C}(\widehat{Y}_R - \widehat{T}) & = \widehat{I}(r,\widehat{Y}) - (\widehat{T} - \widehat{G}) + \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) + \widehat{NI} + \widehat{NT} \\
  \widehat{S}^{pr} & = \widehat{I}(r,\widehat{Y}) - \widehat{S}^{pu} + \widehat{CA} \\
  \widehat{S}^{pr} + \widehat{S}^{pu} - \widehat{I}(r,\widehat{Y}) & = \widehat{CA}
  \end{split}$$
  
 - Where $\widehat{S}^{pr} \equiv (\widehat{Y} + \widehat{NI} + \widehat{NT} - \widehat{T}) - \widehat{C}(\widehat{Y}_R - \widehat{T})$  is the **real** private savings in an open economy, $\widehat{S}^{pu} \equiv \widehat{T} - \widehat{G}$ is the **real** public savings in an open economy and $\widehat{CA} \equiv \widehat{NX}(\widehat{Y}^*,\widehat{Y},\varepsilon) + \widehat{NI} + \widehat{NT}$ is the **current account**.
 
# Acknowledgments

- To my family that supports me

- To the taxpayers of Colombia and the __[UMNG students](https://www.umng.edu.co/estudiante)__ who pay my salary

- To the __[Business Science](https://www.business-science.io/)__ and __[R4DS Online Learning](https://www.rfordatasci.com/)__ communities where I learn __[R](https://www.r-project.org/about.html)__ 
- To the __[R Core Team](https://www.r-project.org/contributors.html)__, the creators of __[RStudio IDE](https://rstudio.com/products/rstudio/)__ and the authors and maintainers of the packages __[tidyverse](https://CRAN.R-project.org/package=tidyverse)__, __[tidyquant](https://CRAN.R-project.org/package=tidyquant)__, __[latex2exp](https://CRAN.R-project.org/package=latex2exp)__,
__[readxl](https://CRAN.R-project.org/package=readxl)__, __[knitr](https://CRAN.R-project.org/package=knitr)__,
__[kableExtra](https://CRAN.R-project.org/package=kableExtra)__ and __[tinytex](https://CRAN.R-project.org/package=tinytex)__ for allowing me to access these tools without paying for a license

- To the __[Linux kernel community](https://www.kernel.org/category/about.html)__ for allowing me the possibility to use some __[Linux distributions](https://static.lwn.net/Distributions/)__ as my main __[OS](https://en.wikipedia.org/wiki/Operating_system)__ without paying for a license

# References
